---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R pacakge zeppr (Beta!)

<!-- badges: start -->
<!-- badges: end -->

The `zeppr` package for R provides a collection of functions to process plant 
protection data. It is actually made for a particular institute but it may be 
useful for more persons. 
It is a beta version provided AS IS without any guaranty or support. 
It is available for non-commercial use at your own risks. 
Commercial users may contact the package developer.


## Installation

This package is available from [GitHub](https://github.com/) and was developed 
using R 4.2.2. The library `devtools` is required for installation. 

You can install a specific release of `zeppr` as follows to get a stable set of 
features:

```{r eval=FALSE}
# uncomment to install devtools if necessary:
# install.packages("devtools")

# Install release v0.2.1-beta (change release tag as required):
devtools::install_github("JFK24/zeppr@v0.2.1-beta") 
```

Alternatively, you can install the development version of `zeppr` as follows to get 
potential new features but also new potential bugs:

```{r eval=FALSE}
devtools::install_github("JFK24/zeppr")
```


## Build tools for package developer

```{r}
# load_all()
# document()
# build_readme()
# test()
# check()
```


## Overview

### Documentation

Full documentation of the ZEPPR functions are available from within R. 
On this page, only examples and use cases are shown. 

### Functions

```{r eval=FALSE}
# Already in the latest release
normalized_cumsum                   # normalized cumulative sum
growing_degree_days                 # growing degree days for 1 day
mutate_cumsum_gdd                   # cumulative sum of growing degree-days
read_isip_hourly_weather_data       # reads weather data from an ISIP Excel file (renamed read_isip_weather_data in devel version!)
mutate_isip_weather_with_cumsum_gdd # cumulative sum of growing degree-days for ISIP weather data

# In development:

# Weather data retrieval
get_dwd_stations_info               # retrieve online info on DWD weather stations
closer_dwd_station                  # get closer DWD station from given coordinates
get_dwd_station_data                # retrieve online data of a DWD weather station
copy_isip_weather_data_to_tsv       # copy ISIP weather data from Excel to TSV 

# temperature index (Klimatologische Kenntage)
day_temp_index                      # classify single days by a temperature index
past_day_temp_index                 # count or classify past days by a temperature index
mutate_past_day_temp_indices        # count or classify multiple indices in a data frame

# vector indexing
index_leading_values_if             # index leading values complying to a condition in a numerical vector
n_leading_values_if                 # count leading values complying to a condition in a numerical vector
index_trailing_values_if            # index trailing values complying to a condition in a numerical vector
n_trailing_values_if                # count trailing values complying to a condition in a numerical vector
index_first_value_if                # index the first value complying to a condition in a numerical vector

```


## Simple use cases for the development version

### Use case - Enrich Weather Data

```{r include = FALSE}
library(zeppr)
library(dplyr)
# Path to an example ISIP weather data file
file <- "20221215_isip_hourly_weather_data_export.xlsx"
path <- system.file("extdata", file, package = "zeppr")

# Read first rows and selected columns of the ISIP file
# daily.table <- read_isip_hourly_weather_data(path, returns.daily.data=TRUE)
daily.table <- read_isip_weather_data(path, hourly.to.daily = TRUE)

# Subset the data and edit some rows to create a fake second location
daily.table <- daily.table[1:10, c("location", "date", "Tmin", "Tmax")]
daily.table[1,]$Tmin <- -1
daily.table[1,]$Tmax <- 4
daily.table[6:10,]$location <- "ABCD" 
daily.table[6:10,]$date <- daily.table[1:5,]$date
daily.table[6:10,]$Tmin <- daily.table[6:10,]$Tmin - 1
daily.table[6:10,]$Tmax <- daily.table[6:10,]$Tmax + 5
```

Let us work on a small fake weather data table with 2 locations.
It was created from a real ISIP file and opened by 
function `read_isip_hourly_weather_data(returns.daily.data=TRUE)`.
It is thus compatible with functions dedicated to ISIP weather data (`mutate_isip_*`).

```{r}
library(zeppr)
library(dplyr)

print(daily.table)
```

Then, we add easily a column for the cumulative sum of 
Growing Degree Days (with default parameters) thanks to an ISIP dedicated 
function: `mutate_isip_weather_with_cumsum_gdd()`. 
This function will take care to separate calculations by locations and years 
if necessary.

We can use an other function to classify days by temperature index or to count 
them in the last days: `mutate_past_day_temp_indices()`. 
Contrary to the growing degree days function, this generic 
function does not care about locations or years. Thus, as our data contains 
multiple locations, we must group it before hands.

```{r}
# add cumulative sum of growing degree-days by location and year (new column: cumsum_gdd)
daily.table <- mutate_isip_weather_with_cumsum_gdd(daily.table, daily.data=TRUE)

# we MUST group the data by locations before counting temperature indices 
daily.table <- group_by(daily.table, location)

# classify days by each temperature index (new columns: ice, frost, vegetation, ...)
daily.table <- mutate_past_day_temp_indices(daily.table, date, Tmin, Tmax, no.suffix=TRUE)

# count days by each temperature index in the last 3 days (vegetation_last_3d, ...)
daily.table <- mutate_past_day_temp_indices(daily.table, date, Tmin, Tmax, n.days = 3)

# count longest period of index-matching days in the last 4 days (vegetation_last_4d_lp, ...) 
daily.table <- mutate_past_day_temp_indices(daily.table, date, Tmin, Tmax, n.days = 4, longest_period = TRUE)

# See results for cumsum_gdd and vegetation index (the table still contains results for all indices: ice, frost, ...)
print(daily.table[, c("location", "date", "Tmin", "Tmax", "cumsum_gdd", "vegetation", "vegetation_last_3d", "vegetation_last_4d_lp")])

```


### Use case - Cumulative Sums

Let us create some toy count and weather data:
```{r}
data.table <- data.frame(
  Date=as.Date(c("2022-03-01", "2022-03-02", "2022-03-03", "2022-03-04", "2022-03-05")),
  Count=c(0, 0, 10, 12, 2),
  Tmin=c(4, 6, 11, 9, 10),
  Tmax=c(12, 14, 13, 16, 14)
  )
head(data.table)
```

Thanks to `zeppr`, we create easily the normalized cumulative sum of the counts and the cumulative sum of growing degree-days (5°C|30°C) as follows:
```{r}
library(zeppr)
data.table$norm_cumsum <- normalized_cumsum(data.table$Count)
data.table <- mutate_cumsum_gdd(
  data.table, date=Date, t.min=Tmin, t.max=Tmax, 
  t.ceiling=30, t.base=5, use.floor=FALSE, hourly.data=FALSE)
head(data.table)
```

Finally we can plot the new values with some plotting functions (not from `zeppr`):
```{r}
plot(data.table$cumsum_gdd, 
     data.table$norm_cumsum, 
     type="o", 
     xlab="cumulative sum of growing degree-days", 
     ylab="normalized cumulative sum of counts")
```

### Use case 2 - DWD Weather Data

```{r include=FALSE}
library(zeppr)
library(dplyr)
library(magrittr)

file.name.1 <- "TU_Stundenwerte_Beschreibung_Stationen.txt"
path.1 <- system.file("extdata", file.name.1, package = "zeppr")
stations.info <- read_dwd_stations_info_file(path.1)

file.name.2 <- "stundenwerte_TU_00044_akt.zip"
path.2 <- system.file("extdata", file.name.2, package = "zeppr")
station.data <- read_dwd_station_data_file(path.2)
```

Let us create some toy locations and corresponding geographic coordinates:
```{r}
library(zeppr)
locations <- data.frame(
  loc=c("A", "B", "C"),
  lat=c(49, 50, 52.52),
  lon=c(8, 9, 13.41)
)
```

Thanks to `zeppr`, we retrieve information on weather stations from DWD weather 
service offering data on air temperature as follows:

```{r eval=FALSE}
stations.info <- get_dwd_stations_info("air_temperature", timerange="recent")
```

```{r}
head(stations.info)
```

We can search the id and distance in km of the closest station for given coordinates:
```{r}
closer_dwd_station(51.89, 10.54, stations_table=stations.info, return_string="id")
closer_dwd_station(51.89, 10.54, stations_table=stations.info, return_string="dist")
```

We can also process the table of locations as follows:
```{r}
# With base R
locations$station.id <- closer_dwd_station(locations$lat, locations$lon, 
  stations_table=stations.info, return_string = "id")
locations$station.name <- closer_dwd_station(locations$lat, locations$lon, 
  stations_table=stations.info, return_string = "name")
head(locations)
```

```{r}
# With dplyr and pipes from magrittr (%>%):
locations %>% 
  mutate(station.id=closer_dwd_station(
    lat, lon, stations_table=stations.info)) %>% 
  mutate(station.name=closer_dwd_station(
    lat, lon, stations_table=stations.info, return_string = "name")) %>% 
  mutate(station.distance=closer_dwd_station(
    lat, lon, stations_table=stations.info, return_string = "dist"))
```

Finally, we can get the weather data from a given station as follows:
```{r eval=FALSE}
station.data <- get_dwd_station_data(station_id="00044", category="air_temperature")
```

```{r}
head(station.data)
```


## Simple documentation of the functions

Extensive documentation is integrated in R help pages
(e.g. type `?growing_degree_days` in the R console to get related help page)

### Normalized cumulative sum for vectors

Function `normalized_cumsum()` calculates the cumulative sum normalized to [0,1].

```{r}
# cumulative sum without normalization
normalized_cumsum(c(1,2,3,4,5), normalize = FALSE)
# with normalization to the total sum (as percentage)
normalized_cumsum(c(1,2,3,4,5))
```


### Growing degree-days for vectors

Function `growing_degree_days()` calculates the growing degree-days 
independently for each pair of min and max day temperatures provided as 2 vectors.
Each pair represents implicitly a time point: either 1 day or 1 hour, though
no time data is required for the calculation here. 
The function is developed for daily data but can be used for hourly data with 
a few tricks explained in annexes and automatized in function `mutate_cumsum_gdd()`
described further below. 
Basic Formula: ((max temperature + min temperature)/2) - base temperature

#### For daily data

```{r}
# A simple example of growing degree-days gdd = ((9+20)/2)-10 = 4.5
growing_degree_days(t.min=9, t.max=20, t.ceiling=30, t.base=10, use.floor=FALSE)
# Same example but t.base is used as floor value: gdd = ((10+20)/2)-10 = 5
growing_degree_days(t.min=9, t.max=20, t.ceiling=30, t.base=10, use.floor=TRUE)
# Now with a lower ceiling value that will change t.max:  gdd = ((10+15)/2)-10 = 2.5
growing_degree_days(t.min=9, t.max=20, t.ceiling=15, t.base=10, use.floor=TRUE)
# Processes pairs of min max temperatures defined in 2 vectors of same length
growing_degree_days(c(7, 8, 10), c(12, 14, 15), t.ceiling=30, t.base=10, use.floor=FALSE)
```


### Cumulative sum of growing degree-days for data frames

Function `mutate_cumsum_gdd()` works on a data frame with date and temperature 
columns to return a copy of the data frame with an additional column for the 
cumulative sum of growing degree-days. 
The **input columns are assumed to be complete**:  no missing time points and no 
missing temperature values. 

#### For daily data

```{r}
# given a simple data.frame with date, min and max temperature
daily.table <- data.frame(
  Date=as.Date(c("2022-01-01", "2022-01-02", "2022-01-03")),
  Tmin=c(4, 6, 11),
  Tmax=c(12, 14, 20))
# we add a column containing cumulative sum of growing degree-days as follows:
mutate_cumsum_gdd(daily.table, date=Date, t.min=Tmin, t.max=Tmax, 
                  t.ceiling=30, t.base=5, use.floor=FALSE, hourly.data=FALSE)
# With %>% pipes from magrittr and custom column name:
library(magrittr)
daily.table %>% mutate_cumsum_gdd(Date, Tmin, Tmax, values.to="my_gdd")
```

#### For hourly data

```{r}
# given a simple data.frame with date.time and temperature (no min and max)
hourly.table <- data.frame(
  date.time=as.POSIXct(c("2022-01-01 13:00:00", "2022-01-02 14:00:00", "2022-01-03 15:00:00")),
  temperature=c(12, 14, 20)
)
# we add a column containing cumulative sum of growing degree-days 
# by setting hourly.data to TRUE and reusing hourly temperatures 
# for both t.min and t.max parameters as follows:
mutate_cumsum_gdd(hourly.table, date=date.time, 
                  t.min=temperature, t.max=temperature, 
                  t.ceiling=30, t.base=5, use.floor=FALSE, hourly.data=TRUE)
# Same with %>% pipes from magrittr and custom column names for various parameters:
library(magrittr)
hourly.table %>% 
  mutate_cumsum_gdd(date.time, temperature, temperature, hourly.data=TRUE, values.to="my_gdd_2", t.base = 2) %>% 
  mutate_cumsum_gdd(date.time, temperature, temperature, hourly.data=TRUE, values.to="my_gdd_3", t.base = 3) %>% 
  mutate_cumsum_gdd(date.time, temperature, temperature, hourly.data=TRUE, values.to="my_gdd_4", t.base = 4)
  
```


### Read ISIP weather data

ISIP weather data files are excel files with multiple sheets (one per 
geographical location) containing each a table of hourly or daily weather data 
for the same date and time range. 
Function `read_isip_weather_data()` reads an ISIP weather data 
file and returns a data frame with either hourly or daily data.
An example file is provided with this package at the following internal path: 

```{r}
file.name <- "20221215_isip_hourly_weather_data_export.xlsx"
path <- system.file("extdata", file.name, package = "zeppr")
```


#### Read the ISIP hourly data file

```{r}
# Get hourly data:
hourly.table <- read_isip_weather_data(path)
head(hourly.table)
```


#### Read the ISIP hourly data file but transform it as daily data

```{r}
# Get daily data:
daily.table <- read_isip_weather_data(path, hourly.to.daily = TRUE)
head(daily.table)
```


### Process ISIP weather data

#### Add the cumulative sum of growing degree-days to ISIP tables

Function `mutate_isip_weather_with_cumsum_gdd()` adds the cumulative sum 
of growing degree-days to an ISIP data frame. Calculations are done on groups 
of rows defined by geographical location and year.

Processing hourly data:

```{r}
# reads file and keeps only 5 first columns and 6 rows (6 hours)
hourly.table <- read_isip_weather_data(path)[1:6,1:5]
# adds cumulative growing degree-days per hour
mutate_isip_weather_with_cumsum_gdd(hourly.table)
```

Processing daily data:

```{r}
# reads file and keeps only 5 first columns and 6 rows (6 days)
daily.table <- read_isip_weather_data(path, hourly.to.daily = TRUE)[1:6,1:5]
# adds cumulative growing degree-days per day
mutate_isip_weather_with_cumsum_gdd(daily.table, daily.data=TRUE)
# adds cumulative growing degree-days per day with floor value for
# min and max temperatures
mutate_isip_weather_with_cumsum_gdd(daily.table, daily.data=TRUE, use.floor=TRUE)
```


## Annexes

### Growing Degree-Days

#### Use the growing degree days daily-data formula to process hourly data 

* Hourly data tables have 1 row per hour and thus 24 rows per day
* Each hour has only 1 temperature (no min and max day temperatures)
* Let us define growing degree-hours as the same as growing degree-days but 
applied for a single hour where the min and max temperatures are both equal to 
the temperature for this hour
* Then, the degree-days for 1 day is equal to the sum of its 24 degree-hours 
  divided by 24. 

The following tricks are automatized by the `mutate_cumsum_gdd()` function 
described further below.
  
```{r}
# Let say we have 24 temperatures for a particular day
temperatures = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 11, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1, 0)
# We can get corresponding degree-hours above 5 degrees
gdh <- growing_degree_days(t.min=temperatures, t.max=temperatures, t.base=5)
print(gdh)
# the growing degree-days for this day is calculated as follows:
sum(gdh)/24
```


